---
title: "IDS 702: Module 1.1"
subtitle: "Motivating example"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
library(knitr)
library(kableExtra)
library(lattice)
library(ggplot2)
library(ggcorrplot)
library(GGally)
```



## Introduction

- By now, you should already be familiar with t-tests and simple linear regression (SLR). 

--

- At the very least, you should know the basics.

--

- Specifically, you should know how to fit a SLR model and assess whether or not the model assumptions are violated. 

--

- We will use those ideas as building blocks for the models we will explore throughout this course.



---
## Motivating example

- In the 1970’s, Harris Trust and Savings Bank was sued for discrimination on the basis of sex.  

--

- As evidence, the defense presented analysis of salaries of employees of one type (skilled, entry level clerical).

--

- The data is in the file `wagediscrim.txt` on Sakai.

--

- We are interested in answering the question: **did female employees tend to receive lower base/starting salaries than similarly qualified and experienced male employees?**

--
<div class="question">
Which statistical tests can we use to probe the question above?
</div>


---
## Data

93 employees on data file (61 female, 32 male).

Variable    | Description
:------------- | :------------ 
bsal | Annual salary at time of hire
sal77 | Annual salary in 1977.
educ | years of education.
exper | months previous work prior to hire at bank.
fsex | 1 if female, 0 if male
senior | months worked at bank since hired
age | months

Since we care about inference on .hlight[bsal], as our response variable, we will exclude .hlight[sal77] for all analysis.

--
<div class="question">
Is this reasonable? Why or why not?
</div>


---
## Data

How many rows? How many columns?
```{r}
wages <- read.csv("data/wagediscrim.txt", header= T)
dim(wages)
```

Take a look at the first few rows of the data.
```{r}
head(wages)
```



---
## Data

Check variable types.
```{r}
wages$sex <- factor(wages$sex,levels=c("Male","Female"))
wages$fsex <- factor(wages$fsex)
str(wages)
```


---
## Exploratory data analysis (EDA)

Next, quick summaries of each variable.
```{r}
summary(wages)
```


---
## EDA

Since we only care about comparing starting salaries for male and female employees for now, let's look at boxplots of .hlight[bsal] by .hlight[sex].
```{r fig.height=2.4}
ggplot(wages,aes(x=sex, y=bsal, fill=sex)) +
  geom_boxplot() + coord_flip() +
  scale_fill_brewer(palette="Blues") +
  labs(title="Baseline Salary vs Sex",y="Base Salary",x="Sex") + 
   theme_classic() + theme(legend.position="none")
```

--
<div class="question">
What do you think? What can you infer from this plot?
</div>


---
## T-test?

We could go further and try a t-test for the hypotheses.
.small[
$$H_0: \mu_{\textrm{male}} - \mu_{\textrm{female}} \leq 0 \ \ \textrm{vs.} \ \ H_A: \mu_{\textrm{male}} - \mu_{\textrm{female}} > 0$$
```{r}
t.test(bsal~sex,data=wages,alternative="greater")
```
]

--
<div class="question">
Is a t-test sufficient here? Any concerns?
</div>



---
## SLR?

How about fitting a SLR model to the two variables.
.block[
.small[
$$ \textrm{bsal}_i = \beta_0 + \beta_1 \textrm{sex}_i + \epsilon_i; \ \ \epsilon_i \overset{iid}{\sim} \mathcal{N}(0, \sigma^2), \ \ \ i = 1,\ldots, n.$$
]
]
```{r}
model1 <- lm(bsal~sex,data=wages); summary(model1)
```

--
<div class="question">
What can we infer from these results?
</div>


---
## EDA

- .block[
T-test shows men started at higher salaries than women $(t=5.83, p < .0001)$; same conclusion from the regression.
]

- But one could argue this is so because both methods **do not** control for other characteristics. Indeed, we have ignored the other variables.

- There are other variables that are correlated with .hlight[bsal]. Here's the correlation matrix of all numerical variables using the .hlight[corr] function in R.
```{r echo=FALSE}
kable(round(cor(wages[,!is.element(colnames(wages),c("sex","fsex"))]),2)) %>%
  kable_styling(font_size = 20)
```



---
## EDA

Or visually (using the `ggcorrplot` package),
```{r fig.height=3.7, warning=F, message=F}
wages_corr <- round(cor(wages[,!is.element(colnames(wages),c("sex","fsex"))]),2)
ggcorrplot(wages_corr, method = "circle",type = "lower",
           colors = c("#6D9EC1", "white", "#E46726"))
```



---
## EDA

- So, let's take a look at scatter plots of all variables

- First, recall the description of all the variables.

Variable    | Description
:------------- | :------------ 
bsal | Annual salary at time of hire
sal77 | Annual salary in 1977.
educ | years of education.
exper | months previous work prior to hire at bank.
fsex | 1 if female, 0 if male
senior | months worked at bank since hired
age | months


---
## EDA

```{r fig.height=3.8}
ggpairs(wages[,!is.element(colnames(wages),c("sal77","sex","fsex"))],
        mapping=ggplot2::aes(colour = "red4",alpha=0.6)) #GGally package
```

This plot looks very busy! 


---
## EDA

Let's take a closer look one variable at a time. First, .hlight[bsal] vs. .hlight[senior].
```{r fig.height=3.3,message=F}
ggplot(wages,aes(x=senior, y=bsal)) +
  geom_point(alpha = .7,aes(color=sex)) +
  geom_smooth(method="lm",col="red3") + theme_classic() +
  labs(title="Baseline Salary vs Seniority",x="Seniority",y="Base Salary")
```


---
## EDA

Next, .hlight[bsal] vs. .hlight[age]
```{r fig.height=3.3,message=F}
ggplot(wages,aes(x=age, y=bsal)) +
  geom_point(alpha = .7,aes(color=sex)) +
  geom_smooth(method="lm",col="red3") + theme_classic() +
  labs(title="Baseline Salary vs Age",x="Age",y="Base Salary")
```


---
## EDA

.hlight[bsal] vs. .hlight[educ]
```{r fig.height=3.3,message=F}
ggplot(wages,aes(x=educ, y=bsal)) +
  geom_point(alpha = .7,aes(color=sex)) +
  geom_smooth(method="lm",col="red3") + theme_classic() +
  labs(title="Baseline Salary vs Education",x="Education",y="Base Salary")
```


---
## EDA

Finally, .hlight[bsal] vs. .hlight[exper]
```{r fig.height=3.3,message=F}
ggplot(wages,aes(x=exper, y=bsal)) +
  geom_point(alpha = .7,aes(color=sex)) +
  geom_smooth(method="lm",col="red3") + theme_classic() +
  labs(title="Baseline Salary vs Experience",x="Experience",y="Base Salary")
```


---
## Takeaways

- Clearly, they are other variables that may be relevant in explaining baseline salary.

--

- We need to explore other statistical methods than the t-test and simple linear regression.

--

- We need methods that can explore the relationship between baseline salary and sex while also controlling for the other variables that clearly may be relevant.

--

- This brings us to .hlight[multiple linear regression (MLR)].

--

- .block[
Something to keep in mind, the overall conclusions may not change after using a better model for this data.

In general, this should never stop you from exploring and reporting the results from better models; you should always be rigorous when doing analyses and be honest when reporting the results!
]




---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




