---
title: "IDS 702: Module 2.1"
subtitle: "Odds, odds ratios, and relative risks"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
library(MASS)
library(arm)
```



## Introduction

- So far, our response variables have been continuous.

--

- Sometimes, we would also like to build models for binary outcome variables. For example,

  + $Y = 1$: healthy, $Y = 0$: not healthy
  + $Y = 1$: employed, $Y = 0$: not employed
  + $Y = 1$: win, $Y = 0$: lose

--

- Often, we want to predict or explain the binary outcome variable from several predictors.

--

- Linear regression is NOT appropriate, because normality for the response variable (and errors) makes no sense in this case.

--

- This brings us to .hlight[logistic regression], the most popular model for binary outcomes.

--

- First let's review relative risk, odds and odds ratios.



---
## Absolute risk and relative risk

- $Y$: binary response variable, $X$: binary predictor

<br />       | $Y=1$       | $Y=0$       |
:----------- | :---------: | :---------: |
$X=1$        | a           | b           |
$X=0$        | c           | d           |

- .hlight[Absolute risk] of $Y=1$ for level $X=1$: $\dfrac{a}{(a+b)}$

- .hlight[Absolute risk] of $Y=1$ for level $X=0$: $\dfrac{c}{(c+d)}$

- .hlight[Relative risk (RR)]: $\dfrac{a/(a+b)}{c/(c+d)}$

--

- Relative risk is a ratio of two probabilities.

--
<div class="question">
Give an example of an application where you think relative risk might be useful.
</div>



---
## Odds and odds ratio

- $Y$: binary response variable, $X$: binary predictor

<br />       | $Y=1$       | $Y=0$       |
:----------- | :---------: | :---------: |
$X=1$        | a           | b           |
$X=0$        | c           | d           |

- .hlight[Odds] of $Y=1$ for level $X=1$: $\dfrac{a}{b}$

- .hlight[Odds] of $Y=1$ for level $X=0$: $\dfrac{c}{d}$

- .hlight[Odds ratio (OR)]: $\dfrac{a/b}{c/d}$

--

- Odds ratio is a ratio of two odds.

--
<div class="question">
Give an example of an application where you think odds or odds ratio might be useful.
</div>



---
## Probabilities and odds: motivating example

- Physicians' Health Study (1989): randomized experiment with 22071 male physicians at least 40 years old.

--

- Half the subjects were assigned to take aspirin every other day.

--

- The other half were assigned to take a placebo pill.

--

- Broad goal: determine whether aspirin decreases cardiovascular mortality.

--

- Here are the number of people in each cell of the contingency table:

<br />         | Heart attack  | No heart attack |
:------------- | :-----------: | :-------------: |
Aspirin        | 104           | 10933           |
Placebo        | 189           | 10845           |



---
## Absolute risk and relative risk for physicians health study

- Physicians Health Study

<br />         | Heart attack  | No heart attack |
:------------- | :-----------: | :-------------: |
Aspirin        | 104           | 10933           |
Placebo        | 189           | 10845           |

- .block[Relative risk of a heart attack when taking aspirin versus when taking a placebo equals]
$$\textrm{RR} = \dfrac{104/(104+10933)}{189/(189+10845)} = 0.55$$

--

- .block[Odds of having a heart attack when taking aspirin over odds of a heart attach when talking a placebo (odds ratio)]
$$\textrm{OR} = \dfrac{104/10933}{189/10845} = 0.546$$



---
## Interpreting odds ratios and relative risks

  <br />       | $Y=1$       | $Y=0$       |
:----------- | :---------: | :---------: |
$X=1$        | a           | b           |
$X=0$        | c           | d           |

- When the variables $X$ and $Y$ are independent 
$$OR = 1; \ \ \ \ \ \ \ \ \ RR = 1$$
  
--

- When subjects with level $X=1$ are more likely to have $Y=1$ than subjects with level $X=0$, then 
$$OR > 1; \ \ \ \ \ \ \ \ \ RR > 1$$
  
--

- When subjects with level $X=1$ are less likely to have $Y=1$ than subjects with level $X=0$, then 
$$OR < 1; \ \ \ \ \ \ \ \ \ RR < 1$$


---
## Relative risk vs. absolute risk: smoking and lung cancer

- Small or large values of relative risk may or may not be significant depending on the base rate.

--

- Thus, it can be more helpful or meaningful to present both the absolute risk and RR.

--

- For example,
  + Percentage of smokers who get lung cancer: 8% (conservative estimate)
  + Relative risk of lung cancer for smokers: 800%
  + That is, getting lung cancer is not commonplace, even for smokers but, smokers’chances of getting lung cancer are much, much higher than non-smokers’ chances.
  + The absolute risk helps place the RR in context.


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




