---
title: "IDS 702: Module 2.3"
subtitle: "Logistic regression with one predictor (illustration)"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(ggplot2)
library(stargazer)
```



## Predicting nba wins

- Let's fit a logistic regression with one predictor to NBA data for four seasons from the 2014/2015 season to the 2017/2018 season.

--

- Suppose we want to see how the amount of points a team let's the opponents score, affects their odds of winning.

--

- For this simple example, we will focus on data from one team: SAS (San Antonio Spurs).

--

- The data is in the file `nba_games_stats_reduced.csv` on Sakai.

--

- Ideally, we should use more information (and that data is actually available) to predict wins but let's continue for illustrative purposes.

--

- You will get to practice with the full data soon.


---
## Predicting nba wins

```{r fig.height=3}
nba <- read.csv("data/nba_games_stats_reduced.csv",header=T)
nba <- nba[nba$Team=="SAS",]
colnames(nba)[3] <- "Opp"
nba$win <- rep(0,nrow(nba)); nba$win[nba$WINorLOSS=="W"] <- 1
nba$win <- as.factor(nba$win)
head(nba); dim(nba)
summary(nba)
```


---
## Predicting nba wins

Only one predictor so not much to do in terms of EDA. We can look at
```{r fig.height=3.4}
ggplot(nba,aes(x=win, y=Opp, fill=win)) +
  geom_boxplot() + coord_flip() +
  scale_fill_brewer(palette="Blues") +
  labs(title="Opposition Points against SAS vs Winning",y="Opposition Points",x="Win") + 
   theme_classic() + theme(legend.position="none")
```








---
## Predicting nba wins

.block[
.small[
$$
\textrm{win}_i | \textrm{Opp}_i \sim \textrm{Bernoulli}(\pi_i); \ \ \ \textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) = \beta_0 + \beta_1 \textrm{Opp}_i
$$
]
]

```{r fig.height=3}
nbareg <- glm(win~Opp,family=binomial(link=logit),data=nba); summary(nbareg)
```

  

---
## Predicting nba wins

Same output re-presented:

```{r fig.height=3,results ='asis',message = FALSE}
stargazer(nbareg,type = "html", header = FALSE,single.row = TRUE)
```

  .block[For every additional point an opponent scores against SAS in a game, the odds of winning decreases by approximately 12%, since exp(-0.126) = 0.88.]



---
## Predicting nba wins

```{r fig.height=3,results ='asis',message = FALSE}
#Let's mean-center the opposition points for interpretation.
nba$Opp_cent <- nba$Opp - mean(nba$Opp)
nbareg <- glm(win~Opp_cent,family=binomial(link=logit),data=nba)
stargazer(nbareg,type = "html", header = FALSE,single.row = TRUE)
```


  .block[The odds of SAS winning an nba game during this period, when the opposition scores approximately 97 points, is approximately 3.11, that is, exp(1.134).]


---
## Predicting nba wins

Confidence intervals for the coefficients. Remember that this is on the log-odds scale.
```{r fig.height=3.4}
confint.default(nbareg) #Asymptotic
confint(nbareg) #Based on the profile-likelihood
```

--
<div class="question">
Can you interpret the intervals?
</div>

---
## Predicting nba wins

Let's transform to the odds scale.
```{r fig.height=3.4}
exp(confint.default(nbareg)) #Asymptotic
exp(confint(nbareg)) #Based on the profile-likelihood
```

--
<div class="question">
Can you interpret the intervals?
</div>



---
## Predicting nba wins

We can get the predicted probabilities for the observed cases.

```{r fig.height=3.4}
nba$predprobs <- predict(nbareg,type="response")
#use predict(logreg, type="link") for the logit scale
nba[1:20,]
```


---
## Predicting nba wins

Useful to examine a plot of predicted probabilities by $x$, that is, opposition points.

```{r fig.height=3.4,message=F}
ggplot(nba,aes(x=Opp, y=predprobs)) +
  geom_point(alpha = .5,colour="blue2") +
  geom_smooth(col="red3") + theme_classic() +
  labs(title="Predicted Prob. of Winning vs Opposition Points",x="Opposition Points",y="Pred. Probability of Winning")
```



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




