---
title: "IDS 702: Module 2.5"
subtitle: "Logistic regression with multiple predictors I"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```



## Logistic regression with multiple predictors: motivating example

- In many developing countries, people get their drinking water from wells.

--

- Sometimes these wells are contaminated with the chemical arsenic, which when consumed in high concentrations causes skin and bladder cancer, as well as cardiovascular disease.

--

- Fortunately, in many cases people living near contaminated wells have the opportunity to get water from nearby uncontaminated wells.



---
## The contaminated wells analysis

- In one study, several researchers measured the concentrations of arsenic in wells in a particular region of Bangladesh.

--

- They labeled wells as safe or unsafe based on the measurements.

--

- The researchers encouraged people drinking from unsafe wells to switch to safe wells.

--

- Several years later, the researchers returned to the area with the goal of seeing who had switched from unsafe to safe wells.

--

- They recorded information on a sample of 3020 individuals who had wells at their homes that were unsafe.

--

- Let's address the question: what predicts why people switch wells?

--

- The data is in the file `arsenic.csv` on Sakai.



---
## The contaminated wells analysis

Data description

Variable    | Description
:------------- | :------------
Switch | 1 = if respondent switched to a safe well <br /> 0 = if still using own unsafe well
Arsenic | amount of arsenic in well at respondent's home (100s of micro-grams per liter)
Dist | distance in meters to the nearest known safe well
Assoc | 1 = if any members of household are active in community organizations <br /> 0 = otherwise
Educ | years of schooling of the head of household

Treat `switch` as the response variable and others as potential predictors.



---
## Logistic regression with multiple predictors

- We can then formally extend the .hlight[logistic regression model] we had before to allow for multiple predictors.

--

- We still have
.block[
.small[
$$\Pr[y_i = 1 | x_i]  = \pi_i \ \ \textrm{and} \ \ \Pr[y_i = 0 | x_i] = 1-\pi_i,$$
]
]

--

  or
  .block[
.small[
$$y_i | x_i \sim \textrm{Bernoulli}(\pi_i)$$
]
]

--

  as before, but with
  .block[
.small[
$$\textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \ldots + \beta_p x_{ip}$$
]
]

  now in both cases.

--

- Let's fit the model to our motivating example.


---
## The contaminated wells analysis: EDA

```{r fig.height=3.4}
arsenic <- read.csv("data/arsenic.csv",header=T,
                    colClasses=c("numeric","numeric","numeric","factor","numeric"))
head(arsenic)
summary(arsenic[,-1])
table(arsenic$switch)
```



---
class: center, middle

# Move to the R script [here](https://ids702-f21.olanrewajuakande.com/slides/Arsenic-I.R).



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




