---
title: "IDS 702: Module 3.6"
subtitle: "Proportional odds model (illustration)"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
library(MASS)
library(arm)
library(e1071)
library(caret)
library(pROC)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```


## Political ideology data

- This data is from a General Social Survey.

--

- Again, political ideology has a five-point ordinal scale, ranging from very liberal to very conservative.

.mini[
<table>
  <tr>
    <th> </th>
    <th> </th>
    <th colspan="5">Political Ideology</th>
  </tr>
  <tr>
    <th colspan="2"></th>
    <td style="text-align:center">Very Liberal</td>
    <td style="text-align:center">Slightly Liberal</td>
    <td style="text-align:center">Moderate</td>
    <td style="text-align:center">Slightly Conservative</td>
    <td style="text-align:center">Very Conservative</td>
  </tr>
  <tr>
    <th rowspan="2">Female</th>
    <td height="50px">Democratic</td>
    <td style="text-align:center">44</td>
    <td style="text-align:center">47</td>
    <td style="text-align:center">118</td>
    <td style="text-align:center">23</td>
    <td style="text-align:center">32</td>
  </tr>
  <tr>
    <td height="50px">Republican</td>
    <td style="text-align:center">18</td>
    <td style="text-align:center">28</td>
    <td style="text-align:center">86</td>
    <td style="text-align:center">39</td>
    <td style="text-align:center">48</td>
  </tr>
  <tr>
    <th rowspan="2">Male</th>
    <td height="50px">Democratic</td>
    <td style="text-align:center">36</td>
    <td style="text-align:center">34</td>
    <td style="text-align:center">53</td>
    <td style="text-align:center">18</td>
    <td style="text-align:center">23</td>
  </tr>
  <tr>
    <td height="50px">Republican</td>
    <td style="text-align:center">12</td>
    <td style="text-align:center">18</td>
    <td style="text-align:center">62</td>
    <td style="text-align:center">45</td>
    <td style="text-align:center">51</td>
  </tr>
</table>
]
  



---
class: center, middle

# In-class analysis: move to the R script [here](https://ids702-f21.olanrewajuakande.com/slides/Political.R)


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!



