---
title: "IDS 702: Module 4.3"
subtitle: "Multilevel/hierarchical linear models (illustration I)"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
library(dplyr)
library(ggplot2)
library(lme4)
#library(lmerTest)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```



## The radon analysis

There are 919 total observations in the data. The data is in the file `Radon.txt` on Sakai.

Variable    | Description
:------------- | :------------
radon | radon levels for each house
log_radon | log(radon)
state | state
floor | lowest living area of each house: 0 for basement, 1 for first floor
countyname | county names
countyID | ID for the county names (1-85)
fips | state + county fips code
uranium | county-level soil uranium
log_uranium | log(uranium)


---
## The radon analysis

```{r fig.height=4,message=F}
Radon <- read.csv("data/Radon.txt", header = T,sep="")
Radon$floor <- factor(Radon$floor,levels=c(0,1),labels=c("Basement","First Floor"))
str(Radon)
head(Radon)
summary(Radon[,-c(2,7)])
```


---
## The radon analysis

.midsmall[
```{r message=F}
table(Radon$countyname) #we don't have enough data in some counties, so we should look to borrow information across counties.
```
]


---
## The radon analysis

The raw radon levels can only take on positive values.

```{r fig.height=3.2,message=F}
ggplot(Radon,aes(radon)) +
  geom_histogram(aes(y=..density..),color="black",linetype="dashed",
                 fill=rainbow(15),bins=15) + theme(legend.position="none") +
  geom_density(alpha=.25, fill="lightblue") + scale_fill_brewer(palette="Blues") +
  labs(title="Distribution of Radon Levels",y="Radon") + theme_classic() 
```

--
.block[Obviously very skewed.]


---
## The radon analysis

Let's look at `log_radon` instead.

```{r fig.height=3.2,message=F}
ggplot(Radon,aes(log_radon)) +
  geom_histogram(aes(y=..density..),color="black",linetype="dashed",
                 fill=rainbow(15),bins=15) + theme(legend.position="none") +
  geom_density(alpha=.25, fill="lightblue") + scale_fill_brewer(palette="Blues") +
  labs(title="Distribution of Log Radon Levels",y="Log Radon") + theme_classic() 
```

--

.block[Much better! Let's go with log radon for now.]



---
## The radon analysis

Are there any variations of radon levels by county? There are too many counties, so, let's do it for a random sample of counties.
```{r fig.height=3.5,message=F,eval=F}
set.seed(1000)
sample_county <- sample(unique(Radon$countyname),25,replace=F)
ggplot(Radon[is.element(Radon$countyname,sample_county),],
       aes(x=countyname, y=log_radon, fill=countyname)) +
  geom_boxplot() +
  labs(title="Log radon levels by county",
       x="County",y="Log Radon") + theme_classic() + 
  theme(legend.position="none",axis.text.x = element_text(angle = 90))
```


---
## The radon analysis

```{r fig.height=4.5,message=F,echo=F}
set.seed(1000)
sample_county <- sample(unique(Radon$countyname),30,replace=F)
ggplot(Radon[is.element(Radon$countyname,sample_county),],
       aes(x=countyname, y=log_radon, fill=countyname)) +
  geom_boxplot() +
  labs(title="Log radon levels by county",
       x="County",y="Log Radon") + theme_classic() + 
  theme(legend.position="none",axis.text.x = element_text(angle = 90))
```

--

.block[Looks like the levels vary by county. However, there are many counties with very little data.]



---
## The radon analysis

Let's focus on counties with at least 11 houses.

```{r fig.height=3.2,message=F,eval=F}
sample_county <- which(table(Radon$countyID) > 10)
ggplot(Radon[is.element(Radon$countyID,sample_county),],
       aes(x=countyname, y=log_radon, fill=countyname)) +
  geom_boxplot() +
  labs(title="Log radon levels by county",
       x="County",y="Log Radon") + theme_classic() + 
  theme(legend.position="none",axis.text.x = element_text(angle = 90))
```



---
## The radon analysis

```{r fig.height=4.5,message=F,echo=F}
sample_county <- which(table(Radon$countyID) > 10)
ggplot(Radon[is.element(Radon$countyID,sample_county),],
       aes(x=countyname, y=log_radon, fill=countyname)) +
  geom_boxplot() +
  labs(title="Log radon levels by county",
       x="County",y="Log Radon") + theme_classic() + 
  theme(legend.position="none",axis.text.x = element_text(angle = 90))
```

<div class="question">
What can you conclude from this plot?
</div>



---
## The radon analysis

Next, the relationship with `floor`, the only individual-level (different observation for each house) variable we have.

```{r fig.height=3,message=F}
ggplot(Radon,aes(x=floor, y=log_radon, fill=floor)) +
  geom_boxplot() + scale_fill_brewer(palette="Greens") +
  labs(title="Log radon vs floor", x="Lowest living area of each house",y="Log Radon") + 
  theme_classic() + theme(legend.position="none")
```

--

.block[Looks like radon levels are higher for houses with the basement as the lowest living area.]

---
## The radon analysis

Let's look at the same relationship for a random sample of counties.

```{r fig.height=3.5,message=F,eval=F}
sample_county <- sample(unique(Radon$countyname),8,replace=F)
ggplot(Radon[is.element(Radon$countyname,sample_county),],
       aes(x=floor, y=log_radon, fill=floor)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Greens") +
  labs(title="Log radon vs floor by county",
       x="Lowest living area of each house",y="Log Radon") + 
  theme_classic() + theme(legend.position="none") +
  facet_wrap( ~ countyname,ncol=4)
```


---
## The radon analysis

```{r fig.height=4.5,message=F,echo=F}
sample_county <- sample(unique(Radon$countyname),8,replace=F)
ggplot(Radon[is.element(Radon$countyname,sample_county),],
       aes(x=floor, y=log_radon, fill=floor)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Greens") +
  labs(title="Log radon vs floor by county",
       x="Lowest living area of each house",y="Log Radon") + 
  theme_classic() + theme(legend.position="none") +
  facet_wrap( ~ countyname,ncol=4)
```

Again, not enough data for some counties.


---
## The radon analysis

Let's focus on counties with at least 16 houses.

```{r fig.height=3.2,message=F,eval=F}
sample_county <- which(table(Radon$countyID) > 15)
ggplot(Radon[is.element(Radon$countyID,sample_county),],
       aes(x=floor, y=log_radon, fill=floor)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Greens") +
  labs(title="Log radon vs floor by county",
       x="Lowest living area of each house",y="Log Radon") + 
  theme_classic() + theme(legend.position="none") +
  facet_wrap( ~ countyname,ncol=4)
```


---
## The radon analysis

```{r fig.height=4.3,message=F,echo=F}
sample_county <- which(table(Radon$countyID) > 15)
ggplot(Radon[is.element(Radon$countyID,sample_county),],
       aes(x=floor, y=log_radon, fill=floor)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Greens") +
  labs(title="Log radon vs floor by county",
       x="Lowest living area of each house",y="Log Radon") + 
  theme_classic() + theme(legend.position="none") +
  facet_wrap( ~ countyname,ncol=4)
```

--

.block[Even though the overall direction is the same, it looks like the actual differences between floor = 0 and floor = 1 differs for some counties.]


---
## The radon analysis

- Let's start by only focusing on `floor`. 

--

- We will try a  varying-slope, varying-intercept linear model.

--

- Let $y_{ij}$ and $x_{1ij}$ be the log radon level and indicator variable `floor` respectively for house $i$ in county $j$.

--

- Mathematically, we have
.block[
.small[
$$
\begin{split}
y_{ij} & = (\beta_{0} + \gamma_{0j}) + (\beta_1 + \gamma_{1j}) x_{1ij} + \epsilon_{ij}; \ \ \ i = 1, \ldots, n_j; \ \ \ j = 1, \ldots, 85\\
\epsilon_{ij} & \sim N(0, \sigma^2) \\
(\gamma_{0j},\gamma_{1j}) & \sim N_2(\boldsymbol{0}, \Sigma).
\end{split}
$$
]
]

--

- Alternative representation:
.block[
.small[
$$
\begin{split}
\text{log(radon}_{ij}\text{)} & = (\beta_{0} + \gamma_{0j}) + (\beta_1 + \gamma_{1j}) \text{ floor}_{ij} + \epsilon_{ij}; \ \ \ i = 1, \ldots, n_j; \ \ \ j = 1, \ldots, 85 \\
\epsilon_{ij} & \sim N(0, \sigma^2) \\
(\gamma_{0j},\gamma_{1j}) & \sim N_2(\boldsymbol{0}, \Sigma).
\end{split}
$$
]
]



---
## The radon analysis

- We skipped this before but $\Sigma$ actually takes the form
.block[
.small[
$$
\Sigma =
\begin{bmatrix}
\tau_0^2 & \rho \tau_0\tau_1  \\
\rho \tau_0\tau_1 & \tau_1^2 \\
\end{bmatrix}
$$
]
]

  where
  + $\tau_0^2$ describes the across county variation attributed to the random/varying intercept,
  
  + $\tau_1^2$ describes the across county variation attributed to the random/varying slope (that is, floor), and
  
  + $\rho$ describes the correlation between $\gamma_{0j}$ and $\gamma_{1j}$.


---
## The radon analysis

In R, we have
```{r fig.height=3.2,message=F}
Model1 <- lmer(log_radon ~ floor + (floor | countyname), data = Radon) 
summary(Model1)
```

---
## Interpretation of fixed effects

- Intuitively, we have an overall "average" regression line for all houses across all counties in Minnesota which has slope -0.68 and intercept 1.46.

--

- That is, the general estimated line for any of the houses in Minnesota is: 
.block[
.small[
$$ \widehat{\text{log(radon}_{i}\text{)}} = 1.46 - 0.68 \times \textrm{floor}_i $$
]
]

- For .hlight[any house in Minnesota with a basement as the lowest living area, the baseline radon level is] $e^{1.46} = 4.31$.

--

- Then, for any house in Minnesota, .hlight[having a first floor as the lowest living area, instead of a basement], reduces the radon level by a multiplicative effect of $e^{-0.68} = 0.51$, that is, about a 49% reduction.

--

- However, if the house is in Dakota county for example, we also need to add on the random intercepts and slopes for that county.



---
## Interpretation of fixed effects

- For Dakota county, we have
  ```{r fig.height=3.2,message=F}
(ranef(Model1)$countyname)["DAKOTA",]
```

  so that the estimated regression line for Dakota county is actually
.block[
.small[
$$ \widehat{\text{log(radon}_{i}\text{)}} = (1.46 - 0.11) + (-0.68-0.09) \times \textrm{floor}_i = 1.35 - 0.77 \times \textrm{floor}_i $$
]
]

--

- Thus, for any house in Dakota county in Minnesota with a basement as the lowest living area, the baseline radon level is actually $e^{1.35} = 3.86$, which is .hlight[lower than the overall state wide average].

--

- And for any house in Dakota county in Minnesota, having the first floor be the lowest living area then reduces the radon level by a multiplicative effect of $e^{-0.77} = 0.46$, that is about a 54% reduction, .hlight[more than the overall state wide effect].


---
## The radon analysis

Again,
```{r fig.height=3.2,message=F}
summary(Model1)
```


---
## Interpretation of random effects

- The estimated standard error $\hat{\sigma} = 0.75$ describes the within-county or remaining unexplained variation.

--

- The estimated $\hat{\tau_0} = 0.35$ describes the across-county variation attributed to the random intercept.

--

- The estimated $\hat{\tau_1} = 0.34$ describes the across-county variation attributed to the random slope (the predictor, floor).

--

- Those two sources of county variation are actually quite similar.

--

- The estimated correlation between $\gamma_{0j}$ and $\gamma_{1j}$ is $\hat{\rho} = -0.34$.

--

- You can visualize the random effects by typing `dotplot(ranef(Model1, condVar=TRUE))$countyname` in R.

--

- So many counties! So, you will need to zoom out on your computer.



---
## Interpretation of random effects

```{r fig.height=5.3,message=F,echo=F}
dotplot(ranef(Model1, condVar=TRUE))$countyname
```


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




