---
title: "IDS 702: Module 5.1"
subtitle: "Introduction to missing data"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
#library(lattice)
#library(dplyr)
#library(ggplot2)
#library(lme4)
#library(arm)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```



## Motivation

- Most real world datasets often suffer from nonresponse, that is, they contain missing values.

--

- Ideally, analysts should first decide on how to deal with missing data before moving on to analysis. 

--

- One needs to make assumptions and ask tons of questions, for example,
  + why are the values missing?
  + what is the pattern of missingness?
  + what is the proportion of missing values in the data?

--

- As a Bayesian, one could treat the missing values as parameters and estimate them simultaneously with the analysis, but even in that case, one must still ask the same questions.

--

- Ask as many questions as possible to help you figure out the most plausible assumptions!


---
## Motivation

- Simplest approach: complete/available case analyses -- delete cases with missing data. Often problematic because:
  + it is just not feasible sometimes (small $n$ large $p$ problem) -- when we have a small number of observations but a large number of variables, we simply can not afford to throw away data, even when the proportion of missing data is small.
  
--

  + information loss -- even when we do not have the small $n$, large $p$ problem, we still lose information when we delete cases.
  
--

  + biased results -- because the missing data mechanism is rarely random, features of the observed data can be completely different from the missing data.

--
  
- More principled approach: impute the missing data (in a statistically proper fashion) and analyze the imputed data.



---
## Why should we care?

- <font color="red">Loss of power</font> due to the the smaller sample size
  + can't regain lost power

--
  
- Any analysis must make an <font color="red">untestable assumption</font> about the missing data
  + wrong assumption $\Rightarrow$ <font color="red">biased estimates</font>

--
  
- Some popular analyses with missing data get <font color="red">biased standard errors</font>
  + resulting in wrong p-values and confidence intervals

--

- Some popular analyses with missing data are <font color="red">inefficient</font>
  + so that confidence intervals are wider than they need be


---
## What to do: loss of power

Approach by design:

- minimize amount of missing data
  + good communications with participants, for example, patients in clinical trial, participants in surveys and censuses, etc
  + follow up as much as possible; make repeated attempts using different methods

--
    
- reduce the impact of missing data
  + collect reasons for missing data
  + collect information predictive of missing values
    
    
---
## What to do: analysis

- A suitable method of analysis would:
  + make the correct (or plausible) assumption about the missing data
  + give an unbiased estimate (under that assumption)
  + give an unbiased standard error (so that p-values and confidence intervals are correct)
  + be efficient (make best use of the available data)

--
  
- However, we can never be sure about what the correct assumption is $\Rightarrow$ sensitivity analyses are essential!


---
## How to approach the analysis?

- Start by knowing:
  + extent of missing data
  + pattern of missing data (e.g. is $X_1$ always missing whenever $X_2$ is also missing?)
  + predictors of missing data and of outcome

--
  
- Principled approach to missing data:
  + identify a plausible assumption (through discussions between you as a data scientist and your clients)
  + choose an analysis method that's valid under that assumption
  
--

- Just because a method is simple to use does not make it plausible; some analysis methods are simple to describe but have complex and/or implausible assumptions.


---
## Types of nonresponse (missing data)

- .hlight[Unit nonresponse]: the individual has no values recorded for any of the variables. For example, when participants do not complete a survey questionnaire at all.

- .hlight[Item nonresponse]: the individual has values recorded for at least one variable, but not all variables.

<table>
  <caption>Unit nonresponse vs item nonresponse</caption>
  <tr>
    <th> </th>
    <th height="30px" colspan="3">Variables</th>
  </tr>
  <tr>
    <th>  </th>
    <td height="30px" style="text-align:center" width="100px"> X<sub>1</sub> </td>
    <td style="text-align:center" width="100px"> X<sub>2</sub> </td>
    <td style="text-align:center" width="100px"> Y </td>
  </tr>
  <tr>
    <td height="30px" style="text-align:left"> Complete cases </td>
    <td style="text-align:center"> &#10003 </td>
    <td style="text-align:center"> &#10003 </td>
    <td style="text-align:center"> &#10003 </td>
  </tr>
  <tr>
    <td rowspan="3"> Item nonresponse </td>
    <td rowspan="3" style="text-align:center"> &#10003 </td>
    <td height="30px" style="text-align:center"> &#10003 </td>
    <td style="text-align:center"> &#10067 </td>
  </tr>
  <tr>
    <td height="30px" style="text-align:center"> &#10067 </td>
    <td style="text-align:center"> &#10067 </td>
  </tr>
  <tr>
    <td height="30px" style="text-align:center"> &#10067 </td>
    <td style="text-align:center"> &#10003 </td>
  </tr>
  <tr>
    <td height="30px" style="text-align:left"> Unit nonresponse </td>
    <td style="text-align:center"> &#10067 </td>
    <td style="text-align:center"> &#10067 </td>
    <td style="text-align:center"> &#10067 </td>
  </tr>
</table>



---
## Types of Missing Data Mechanism

- Data are said to be .hlight[missing completely at random (MCAR)] if the reason for missingness does not depend on the values of the observed data or missing data.

--

- For example, suppose
  - you handed out a double-sided survey questionnaire of 20 questions to a sample of participants;
  - questions 1-15 were on the first page but questions 16-20 were at the back; and
  - some of the participants did not respond to questions 16-20.
 
--
 
- Then, the values for questions 16-20 for those people who did not respond would be .hlight[missing completely at random] if they simply did not realize the pages were double-sided; they had no reason to ignore those questions.
 
--
 
- This is rarely plausible in practice!


---
## Types of Missing Data Mechanism

- Data are said to be .hlight[missing at random (MAR)] if the reason for missingness may depend on the values of the observed data but not the missing data (conditional on the values of the observed data).

--

- Using our previous example, suppose
  - questions 1-15 include demographic information such as age and education;
  - questions 16-20 include income related questions; and
  - once again, some of the participants did not respond to questions 16-20.

--
  
- Then, the values for questions 16-20 for those people who did not respond would be .hlight[missing at random] if younger people are more likely not to respond to those income related questions than old people, where age is observed for all participants.
  
--

- This is the most commonly assumed mechanism in practice!


---
## Types of Missing Data Mechanism

- Data are said to be .hlight[missing not at random (MNAR or NMAR)] if the reason for missingness depends on the actual values of the missing (unobserved) data.

--

- Continuing with our previous example, suppose again that
  - questions 1-15 include demographic information such as age and education;
  - questions 16-20 include income related questions; and
  - once again, some of the participants did not respond to questions 16-20.

--
  
- Then, the values for questions 16-20 for those people who did not respond would be .hlight[missing not at random] if people who earn more money are less likely to respond to those income related questions than old people.

--
  
- This is usually the case in real analysis, but analysis can be complex!


---
## Types of missing data mechanisms: how to tell in practice?

So how can we tell the type of mechanism we are dealing with? 

--

In general, we don't know!!!

--

- Rare that data are MCAR (unless planned beforehand)

--

- Possible that data are MNAR

--

- Compromise: assume data are MAR if we include enough variables in model for the missing data indicator $\boldsymbol{R}$.


---
## Why should we care?

- <div class="question">
Why should we care in practice? What does bias really mean here? How exactly does using only the complete cases affect our results for the three mechanisms?
</div>

- Let's attempt to answer these questions via simulations. 

- Set $n = 10,000$. For $i=1,\ldots,n$, generate
  + $x_i \overset{iid}{\sim} N(2, 1); \ \  \ y_i|x_i \overset{iid}{\sim} N(-1 + 2 x_{i}, \sigma^2=5^2)$
  + $r_i|y_i, x_i \sim \textrm{Bernoulli}(\pi_i); \ \ \ \textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) = \theta_0 + \theta_1 y_i + \theta_2 x_i$
  
- Next, set $y_i$ missing whenever $r_i = 1$.

- Set different values for $\boldsymbol{\theta} = (\theta_0, \theta_1, \theta_2)$ to reflect MCAR, MAR and MNAR.

- Let's use the R script [here](https://ids702-f21.olanrewajuakande.com/slides/MissingDataSim.R).


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




