---
title: "IDS 702: Module 7.1"
subtitle: "Introduction to time series analysis"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
library(ggplot2)
#library(arm)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```



## Introduction

- When data are ordered in time, responses and errors from one period may influence responses and errors from another period.

--

- For example, it is reasonable to expect unemployment rate in a month to be correlated with unemployment rate in previous month(s).

--

- Another example: weather events in current time period may depend on weather events in previous time period.

--

- These are called .hlight[time series] data.

--

- Correlation due to time is called serial correlation or autocorrelation.

--

- We will only scratch the surface in this course.


---
## Goals of time series analysis

- .hlight[Forecasting outcomes]

--

  + Given a series of outcomes ordered in time, predict the values of the outcomes in the future.

--
  
  + Examples:
      - forecasting future price of oil given historical oil prices.
      - predicting future price of a particular stock price given past prices of the same stock.

--

  + When forecasting, it is important to also report an interval estimate to incorporate uncertainty about future values.
  

---
## Goals of time series analysis

- .hlight[Forecasting outcomes]
  
--

  + Forecasting outcomes using predictors may involve building a model for the predictors as well, since we can't observe them in the future.
  
--

  + For example, predicting inflation rate given employment rate requires estimating future values for the employment rate as well.
  
--

- .hlight[Learning relationships with data ordered in time].
  
--

  + How are outcomes correlated over time? Are there periodic relationships in outcomes?
  
--

  + Regressions of outcomes on predictors, accounting for correlated errors due to time series.
  

---
## Motivating example: FTSE 100

- The FTSE (Financial Times Stock Exchange) 100 Index is a share index of the 100 companies listed on the London Stock Exchange with the highest market capitalization.
  
--

- A share index is essentially a form of weighted average of prices of selected stocks.
  
--

- To motivate our discussions on time series, let's look at data for FTSE 100 returns in 2018.
  ```{r}
ftse100 <- read.csv("data/ftse2018.csv", header = T)
head(ftse100)
```

- Can we forecast closing prices for the next five days from 11/7/2018?

---
## Motivating example: FTSE 100

Notice that the data go from latest to earliest date, so let's invert the order of the rows to make the time series increasing in date.
  
--

```{r}
ftse100 <- ftse100[nrow(ftse100):1,]
dim(ftse100)
head(ftse100)
```

---
## Motivating example: FTSE 100

Plot the closing prices to see what a simple time series data looks like.
```{r fig.height=2.8}
tsClose <- ts(ftse100$Close); ts.plot(tsClose,col="red3")
```
  
--

- It is reasonable to expect closing prices for a particular day to be correlated with closing prices for previous days. 
  
--

- How many of the previous days? We will have to investigate!


---
## Motivating example: Sunspots and melanoma

- We will revisit that data but let's look at different example, where we also have a predictor.
  
--

- Incidence of melanoma (skin cancer) may be related to solar radiation.
  
--

- Annual data from Connecticut tumor registry on age adjusted melanoma incidence rates (per 100000 people).

--

- Treat these rates as without error.
  
--

- We also have annual data on relative sunspot (dark spots on the sun caused by intense magnetic activity) activity.
  
--

- Data go from 1936 to 1972.


---
## Motivating example: Sunspots and melanoma

```{r fig.height=3}
cancersun <- read.csv("data/melanoma.csv", header = T)
names(cancersun) = c("year", "melanoma", "sunspot")
str(cancersun)
head(cancersun)
```


---
## Motivating example: Sunspots and melanoma

```{r fig.height=3, message=F}
ggplot(cancersun, aes(x=sunspot, y=melanoma)) +
  geom_point(alpha = .5,colour="blue4") +
  geom_smooth(method="lm",col="red3") +
  labs(title="Melanoma Incidence Rate vs Sunspots") +
  theme_classic()
```

Weak positive (maybe!) relationship between them.


---
## Motivating example: Sunspots and melanoma

Let's look at melanoma incidence rate in time
```{r fig.height=3}
tsMelanoma <- ts(cancersun$melanoma); ts.plot(tsMelanoma,col="blue4")
```

Trend in time, some of which we might be able to explain using `sunspots`.


---
## Motivating example: Sunspots and melanoma

Let's fit a linear model to the relationship between the two variables.
```{r fig.height=2.8 , message=F}
regmelanoma = lm(melanoma ~ sunspot, data = cancersun)
ggplot(cancersun, aes(x=sunspot, y=regmelanoma$residual)) +
  geom_point(alpha = .5,colour="blue4") +
  geom_smooth(method="lm",col="red3") + labs(title="Residuals vs Sunspots") +
  theme_classic()
```

Residuals look fine here.

---
## Motivating example: Sunspots and melanoma

Let's plot the residuals versus year.
```{r fig.height=2.8, message=F}
ggplot(cancersun, aes(x=year, y=regmelanoma$residual)) +
  geom_point(alpha = .5,colour="blue4") +
  geom_smooth(method="lm",col="red3") + labs(title="Residuals vs Year") +
  theme_classic()
```

Huge trend! What to do???


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




