---
title: "IDS 702: Module 7.3"
subtitle: "AR and MA models"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
library(ggplot2)
#library(arm)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```



## AR models

- The most common time series model is called the .hlight[autoregressive (AR)] model.
  
--

- When only one lag matters, the zero-mean AR(1) model is
.block[
.small[
$$y_t = \phi y_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- With a non-zero mean, we have
.block[
.small[
$$y_t = \mu + \phi y_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- When the mean is non-zero, we can choose to de-mean (mean-center) the series and model that instead.
  
--

- In both cases, for the AR(1) we basically have a linear regression where the value of the outcome at time $t$ depends on value of outcome at time $t-1$.
  
--

- $\phi$ is the autocorrelation.


---
## AR models

- For the zero-mean AR(1) model,  

--

  + $|\phi|<1$ represents stationary time series.
    
--

  + $\phi=1$ is a random walk.
    
--

  + $|\phi|>1$ implies non-stationary, "explosive" models.
    
--

  
- A stationary AR(1) series varies around its mean, randomly wandering off away from the mean in response to the "input" values of the random $\epsilon_t$ series, but always returning to near the mean, and never "exploding" away for more than a short time.
  
--

- AR(1) series with $0<\phi<1$ represent short-term, positive correlations that would damp out exponentially if $\epsilon_t$ were zero.
  
--

- Negative values of $\phi$ represent short-term, negative correlations.


---
## AR models

- Let's explore what AR(1) models look like via simulations.
  
--

- Move to the R script [here](https://ids-702-f20.github.io/Course-Website/slides/TS_simulations.R).

--

- Note that
  
--

  + autocorrelations decay steadily with lags.
    
--

  + partial autocorrelations go to zero after lag p.


---
## AR models

- For a zero mean AR(p) model, we have
.block[
.small[
$$y_t = \sum_{k=1}^{p} \phi_k y_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- So that for a non-zero mean AR(p) model, we have
.block[
.small[
$$y_t = \mu + \sum_{k=1}^{p} \phi_k y_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- AR(p) models are capable of adequately representing a wide range of observed behaviors in time series for large enough $p$.


---
## AR models: how many lags?

- Several ways to decide how many lags to include.
  
--

- Use graphical techniques
  + Look at partial autocorrelation plots.
    
--

  + Set $p$ at lag where correlations become small enough not to be important.
  
--

- Use a model selection criterion like BIC.
  
--

- See section 8.6 of the assigned readings.
  
--

- Sometimes in time series data, the partial autocorrelations are small even at lag 1.
  
--

- In this case, it can be reasonable to skip autoregressive models and just use usual linear regression modeling approaches.


---
## What if the series is not stationary?

- Sometimes transformations can make stationarity a reasonable assumption.
  
--

- Differencing (subtract lagged values from outcome at time $t$) also often help; changes over time are more likely to be stationary than the raw values.
  
--

- Including predictors can also help as we will see later with the melanoma example.
  
--

- There are other models for non stationary time series.


---
## AR(p): including predictors

- We also might want to account for serial correlation in regression modeling.
  
--

- Linear regression assumes independent errors across individuals.
  
--

- As we have already seen with the melanoma example, this may not be reasonable with time series data.
  
--

- With a single predictor $x_t$, we have
.block[
.small[
$$y_t = \mu + \sum_{k=1}^{p} \phi_k y_{t-k} + x_t + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- That is, the value of outcome at time $t$ depends on value of outcome at time $t-1, t-2, \ldots, t-k$, but also on the predictor $x$ at time $t$.
  
--

- Easy to extend the model to multiple predictors.


---
## Model assumptions: stationarity

- Coefficients and regression variance do not change with time.
  
--

  + Apart from changes in explanatory variables, the behavior of the time series is the same at different segments of time.
    
--

  + Generally, no predictable patterns in the long term
    
--

- Diagnostics: check if patterns in residuals are similar across time.
  
--

- Tests:
  + Ljung-Box
  + Augmented Dickey–Fuller (ADF)
  + Kwiatkowski-Phillips-Schmidt-Shin (KPSS)
  
--

- Remedies:
  + Sometimes transformations (e.g., using logs) can make stationarity more reasonable.
  + Use time series models that allow for drifts.
  
  
---
## Model assumptions: others

- Other assumptions
  1. Linearity
  
  2. Independence of errors
  
  3. Equal variance

  4. Normality
  
--

- Diagnose using the same methods we used for linear regression.
  
--

- Remedies include transformations and model changes as we had before.


---
## MA models

- The zero-mean MA(1) model is
.block[
.small[
$$y_t = \phi \epsilon_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- With a non-zero mean, we have
.block[
.small[
$$y_t = \mu + \phi \epsilon_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- The value of the outcome at time $t$ depends on the value of the deviation from the mean (the error term) at time $t-1$.
  
--

- For a zero mean MA(p) model, we have
.block[
.small[
$$y_t = \sum_{k=1}^{p} \phi_k \epsilon_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- So that for a non-zero mean MA(p) model, we have
.block[
.small[
$$y_t = \mu + \sum_{k=1}^{p} \phi_k \epsilon_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]


---
## MA models

- Let's explore what MA(1) models looks like via simulations. Move back to the same R script.

--

- Note that
  
--

  + Autocorrelations die off almost immediately after lag 1.
    
--

  + In MA(p) model, autocorrelations (mostly!) die off after lag $p$. May not be exact since autocorrelation measures correlation between the actual outcome at different time points.
    
--

  + Partial autocorrelations are not particularly useful.
  
<!-- - Contrast to AR(p) models -->
<!--   + Autocorrelations tend to decrease over time smoothly. -->
<!--   + Partial autocorrelations die off after lag $p$. -->
    
--

- It is possible to write any stationary AR(p) model as an $\textrm{MA}(\infty)$ model. The reverse result holds for some constraints on the MA parameters. See the reading material.
  
  
---
## Deciding models?

- Use autocorrelations and partial autocorrelations to help decide model.
  
--

- Steady decay on autocorrelations often implies AR.
  
--

- Non zero autocorrelations before lag $p$ and zero after lag $p$ often implies MA.
  
--

- Sometimes use both AR and MA error structure, called an .hlight[ARMA] model.
  
--

- Whenever we take differences in $y$ values to ensure stationarity before fitting ARMA models, we have .hlight[ARIMA] models.



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!



