---
title: "IDS 702"
subtitle: "Instructons for final project presentations"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```




## Final project presentations

- Each student will create/record a 6 minute video presentation of their findings.
  
--

- Findings can be (mostly!) preliminary.
  
--

- Presentations will be graded for:
  + Understanding of the data/project
  
  + Clarity
  
  + Communication of findings

--

- Each student's presentation must be uploaded by 11:59pm on Monday, November 22.

--

- The presentations MUST be uploaded using the VoiceThread tool on Sakai.

--

- Final project reports are due Tuesday, December 12, on Gradescope.



---
## Final project presentations

- To upload your presentation, go to .hlight[Sakai] $\rightarrow$ .hlight[VoiceThread] $\rightarrow$ .hlight[Add your own].

--

- Click .hlight[Create a new VoiceThread] and drop your media file from your local computer.

--

- The title of your file must be written as .hlight[Title of Your Project (Your Full Name)].

--

- Next, click on .hlight[Share].

--

- Under the .hlight[Secure] tab, be sure to select .hlight[IDS.702.01.F21].

--

- Next, click on .hlight[Share] again, and you can close the window once you see .hlight[Success].


---
## Final project presentations

- Finally, each student MUST also watch .hlight[at least three presentations] from the other students.

--

- Browse through the titles to find the topics you find interesting.

--

- You must leave at least one comment under each video you watch using the .hlight[comment button].

--

- Please leave constructive feedback, questions or suggestions for your classmates.

--

- **You will be helping each other a ton by doing so!**

--

- Please make sure to watch the videos by Friday, December 3.
  





