---
title: "IDS 702: Module 1.10"
subtitle: "Bringing the MLR pieces together I (illustration)"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(rvest)
```



## Diamonds data

- A diamond's value is often determined using four factors known as the 4Cs: color, clarity, cut (certification) and carat weight.

--

  + Color: evaluation based on absence of color; how pure the diamond is. .hlight[This is a categorical variable with 6 levels.]
  
--
  
  + Clarity: evaluation based on absence of blemishes. .hlight[This is a categorical variable with 5 levels.]
  
--

  + Certification: how well the diamond is cut; how well a diamond's facets interacts with light. .hlight[This is a categorical variable with 3 levels.]

--

  + Carats: carat weight measuring how much the diamond weighs. .hlight[This is a continuous variable.]
  
--

- We will use some data to draw inference about how these factors affect a diamond's price .hlight[(continuous)].

--

- You can read more about the 4Cs [here](https://4cs.gia.edu/en-us/4cs-diamond-quality/).



---
## Multiple regression of diamonds data

- A good starting model is
.block[
$$y_i = \boldsymbol{x}_i\boldsymbol{\beta} + \epsilon_i; \ \ \epsilon_i \sim N(0, \sigma^2).$$
]

  where $y_i$ is the price for observation $i$, and $\boldsymbol{x}_i$ is the vector containing the corresponding values for Carats, Color, Clarity, and Certification.
  
--

- Alternatively, write
.block[
.midsmall[
$$
\begin{split}
\text{Price}_i & = \beta_0 + \beta_1 \text{Carats}_i + \sum_{j=2}^6 \beta_{2j} \mathbb{1}[\text{Color}_i = j] + \sum_{j=2}^5 \beta_{3j} \mathbb{1}[\text{Clarity}_i = j] \\
& \ \ \ + \sum_{j=2}^3 \beta_{4j} \mathbb{1}[\text{Certification}_i = j] +  \epsilon_i; \ \ \epsilon_i \sim N(0, \sigma^2).
\end{split}
$$
]
]

--

- Can also write
.block[
.midsmall[
$$
\begin{split}
\widehat{\text{Price}}_i & = \hat{\beta}_0 + \hat{\beta}_1 \text{Carats}_i + \sum_{j=2}^6 \hat{\beta}_{2j} \mathbb{1}[\text{Color}_i = j] + \sum_{j=2}^5 \hat{\beta}_{3j} \mathbb{1}[\text{Clarity}_i = j] \\
& \ \ \ + \sum_{j=2}^3 \hat{\beta}_{4j} \mathbb{1}[\text{Certification}_i = j].
\end{split}
$$
]
]




---
## Multiple regression of diamonds data

- This is just a candidate model. 

--

- We will go through the full (almost!) modeling process and we will see if this model makes sense or if we need to make changes to it.

--

- We will start by doing EDA, all the way down to model assessment, including investigating multicollinearity.

--

- We will explore transformations, polynomial forms, interactions, etc.

--

- The data is in the file `diamonds.csv` on Sakai.




---
class: center, middle

# Move to the R script [here](https://ids702-f21.olanrewajuakande.com/slides/Diamonds.R).



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




