---
title: "IDS 702: Module 1.12"
subtitle: "Bringing the MLR pieces together II (illustration)"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
```



## Back to the diamonds data

Let's try model selection for our diamonds example. We will do this on the log scale - recall our analysis in the previous module.

First, forward selection using AIC

```{r,error=FALSE}
diamonds <- read.csv("data/diamonds.csv", header= T,
                     colClasses = c("numeric","factor","factor","factor","numeric"))
diamonds$CaratsCent <- diamonds$Carats - mean(diamonds$Carats)
diamonds$CaratsCent2 <- diamonds$CaratsCent^2
NullModel <- lm(log(Price)~1,data=diamonds)
FullModel <- lm(log(Price)~CaratsCent+CaratsCent2+
                  Color*Clarity+Color*Certification+
                  Clarity*Certification,
                data=diamonds)
Model_forward <- step(NullModel, scope = formula(FullModel),direction="forward",trace=0)
# Remove the trace=0 option if you want to function to print the entire process
# Let's see the variables the model selected
Model_forward$call
#run summary(Model_forward) to see the results of the final model
```


---
## Back to the diamonds data

Let's do the same using BIC

```{r}
# use k = log(n) to use BIC instead.
n <- nrow(diamonds)
Model_forward <- step(NullModel, scope = formula(FullModel),direction="forward",trace=0,
                      k = log(n))
# Let's see the variables the model selected
Model_forward$call
#run summary(Model_forward) to see the results of the final model
```


---
## Back to the diamonds data

Backward selection using AIC

```{r}
Model_backward <- step(FullModel,direction="backward",trace=0)
# Let's see the variables the model selected
Model_backward$call
#run summary(Model_backward) to see the results of the final model
```

Same result as forward selection using AIC


---
## Back to the diamonds data

Backward selection using BIC

```{r}
Model_backward <- step(FullModel,direction="backward",trace=0,k = log(n))
# Let's see the variables the model selected
Model_backward$call
#run summary(Model_backward) to see the results of the final model
```

Same result as forward selection using BIC


---
## Back to the diamonds data

Stepwise selection using AIC

```{r}
Model_stepwise <- step(NullModel, scope = formula(FullModel),direction="both",trace=0)
# Let's see the variables the model selected
Model_stepwise$call
#run summary(Model_backward) to see the results of the final model
```

Same result as previous results using AIC


---
## Back to the diamonds data

Stepwise selection using BIC

```{r}
Model_stepwise <- step(NullModel, scope = formula(FullModel),direction="both",trace=0,
                       k = log(n))
# Let's see the variables the model selected
Model_stepwise$call
#run summary(Model_backward) to see the results of the final model
```

Same result as previous results using BIC


---
## Back to the diamonds data

Let's use the .hlight[regsubsets] function.

```{r}
library(leaps)
Model_forward <- regsubsets(log(Price)~CaratsCent+CaratsCent2+Color*Clarity+
                              Color*Certification+Clarity*Certification,data=diamonds,
                            method="forward")
Select_results <- summary(Model_forward)
coef(Model_forward, which.max(Select_results$adjr2)) # Adj R-sq 
coef(Model_forward, which.min(Select_results$bic)) #BIC
```

---
## Back to the diamonds data

```{r}
Model_backward <- regsubsets(log(Price)~CaratsCent+CaratsCent2+Color*Clarity+
                              Color*Certification+Clarity*Certification,data=diamonds,
                            method="backward")
Select_results <- summary(Model_backward)
coef(Model_backward, which.max(Select_results$adjr2)) # Adj R-sq 
coef(Model_backward, which.min(Select_results$bic)) #BIC
```





---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




