---
title: "IDS 702: Module 2.6"
subtitle: "Logistic regression with multiple predictors II"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```



## The contaminated wells analysis cont'd

Data description:

Variable    | Description
:------------- | :------------
Switch | 1 = if respondent switched to a safe well <br /> 0 = if still using own unsafe well
Arsenic | amount of arsenic in well at respondent's home (100s of micro-grams per liter)
Dist | distance in meters to the nearest known safe well
Assoc | 1 = if any members of household are active in community organizations <br /> 0 = otherwise
Educ | years of schooling of the head of household

--

Treat `switch` as the response variable and others as potential predictors.

--

The data is in the file `arsenic.csv` on Sakai.



---
class: center, middle

# Move to the R script [here](https://ids702-f21.olanrewajuakande.com/slides/Arsenic-II.R).



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




