---
title: "IDS 702: Module 3.4"
subtitle: "Multinomial logistic regression (illustration)"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Analysis of Sesame Street Data

- The television series Sesame Street is concerned mainly with teaching preschool skills to children age 3-5, with special emphasis on reaching economically disadvantaged children.

--

- In the early 1970s, researchers at Educational Testing Service (the company that runs the SAT) ran a study to evaluate Sesame Street.

--

- To ensure the study contained a group of children that watched Sesame Street regularly, they randomly assigned children either to receive encouragement to watch Sesame Street or not to receive encouragement.

--

- Those assigned to encouragement were given promotional materials, and received weekly visits and phone calls from ETS staff.  Those assigned not to receive encouragement did not get this attention.

--

- The children were tested on a variety of cognitive variables, including knowledge of body parts, knowledge about letters, knowledge about numbers, etc., both before and after viewing the series.

--

- .hlight[Let's predict how often the kids watch sesame street, with focus on whether encouragement pushes them towards more viewing.]


---
## Analysis of Sesame Street Data

The data is in the file `sesame.txt` on Sakai.

.mini[
Variable    | Description
:------------- | :------------
viewcat |  1=rarely watched the show <br/> 2=once or twice a week <br/> 3=three to five times a week <br/> 4=watched the show on average more than 5 times a week
viewenc |   1=child encouraged to watch,  2=child not encouraged to watch
site |   1 =Three to five year old disadvantaged children from inner city areas in various parts of the country. <br/> 2 = Four year old advantaged suburban children. <br/> 3 = Advantaged rural children. <br/> 4 = Disadvantaged rural children. <br/> 5 = Disadvantaged Spanish speaking children.
sex  | male=1, female=2
age  | age in months
setting |   setting in which Sesame Street was viewed, 1=home 2=school
prebody |  pretest on knowledge of body parts (scores range from 0-32)
prelet |  pretest on letters (scores range from 0-58)
preform |  pretest on forms (scores range from 0-20)
prenumb | pretest on numbers (scores range from 0-54)
prerelat | pretest on relational terms (scores range from 0-17)
preclasf | pretest on classification skills
]



---
class: center, middle

# In-class analysis: move to the R script [here](https://ids702-f21.olanrewajuakande.com/slides/Sesame.R)


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




