---
title: "IDS 702: Module 4.6"
subtitle: "Multilevel/hierarchical logistic models (illustration)"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
#library(lattice)
#library(dplyr)
#library(ggplot2)
library(lme4)
library(arm)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```


## 1988 elections analysis recap

2193 observations from one of eight CBS News surveys.

.small[
Variable    | Description
:------------- | :------------
org | cbsnyt = CBS/NYT
bush | 1 = preference for Bush Sr., 0 = otherwise
state | 1-51: 50 states including DC (number 9)
edu | education: 1=No HS, 2=HS, 3=Some College, 4=College Grad
age | 1=18-29, 2=30-44, 3=45-64, 4=65+
female | 1=female, 0=male
black | 1=black, 0=otherwise
region | 1=NE, 2=S, 3=N, 4=W, 5=DC
v_prev | average Republican vote share in the three previous elections (adjusted for home-state and home-region effects in the previous elections)
]

The data is in the file `polls_subset.txt` on Sakai.


---
## 1988 elections analysis recap

```{r}
polls_subset <- read.table("data/polls_subset.txt",header=TRUE)
polls_subset$v_prev <- polls_subset$v_prev*100 #rescale 
polls_subset$region_label <- factor(polls_subset$region,levels=1:5,
                                    labels=c("NE","S","N","W","DC"))
polls_subset$edu_label <- factor(polls_subset$edu,levels=1:4,
                                 labels=c("No HS","HS","Some College","College Grad"))
polls_subset$age_label <- factor(polls_subset$age,levels=1:4,
                                 labels=c("18-29","30-44","45-64","65+"))
data(state) 
state_abbr <- c (state.abb[1:8], "DC", state.abb[9:50])
polls_subset$state_label <- factor(polls_subset$state,levels=1:51,labels=state_abbr)
rm(list = ls(pattern = "state"))
```



---
## 1988 elections analysis

- I will not do any substantial EDA here.

--

- I expect you to be able to do this yourself.

--

- Let's just take a look at the amount of data we have for "bush" and the age:edu interaction.

  ```{r}
###### Exploratory data analysis
table(polls_subset$bush) #well split by the two values
table(polls_subset$edu,polls_subset$age)
```


---
## 1988 elections analysis

- As a start, we will consider a simple model with fixed effects of race and sex and a random effect for state (50 states + the District of Columbia).
.block[
.small[
$$
\begin{split}
\textrm{bush}_i | \boldsymbol{x}_i & \sim \textrm{Bernoulli}(\pi_i); \ \ \ i = 1, \ldots, n; \ \ \ j = 1, \ldots, J=51; \\
\textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) & = \beta_{0} + \gamma_{0j[i]} + \beta_1 \textrm{female}_{i} + \beta_2 \textrm{black}_{i}; \\
\gamma_{0j} & \sim N(0, \sigma_{\textrm{state}}^2).
\end{split}
$$
]
]

--

- We can also write
.block[
.small[
$$
\begin{split}
\textrm{bush}_i | \boldsymbol{x}_i & \sim \textrm{Bernoulli}(\pi_i); \ \ \ i = 1, \ldots, n; \ \ \ j = 1, \ldots, J=51; \\
\textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) & = \beta_{0} + \gamma_{0j[i]}^{\textrm{state}} + \beta_{\textrm{female}} \textrm{female}_{i} + \beta_{\textrm{black}} \textrm{black}_{i}; \\
\gamma_{0j} & \sim N(0, \sigma_{\textrm{state}}^2).
\end{split}
$$
]
]

--

- In `R`, we have
  ```{r results = 'hide'}
library(lme4)
model1 <- glmer(bush ~ black+female+(1|state_label),family=binomial(link="logit"),
data=polls_subset)
summary(model1)
```


---
## 1988 elections analysis
```{r echo=F}
summary(model1)
```


---
## 1988 elections analysis

- Looks like we dropped some NAs.
  ```{r}
c(sum(complete.cases(polls_subset)),sum(!complete.cases(polls_subset)))
```

--

- Not ideal; we'll learn about methods for dealing with missing data soon.

--

- Interpretation of results:
  + For a fixed state (or across all states), a non-black male respondent has odds of $e^{0.45}=1.57$ of supporting Bush.
  
--

  + For a fixed state and sex, a black respondent as $e^{-1.74}=0.18$ times (an 82% decrease)  the odds of supporting Bush as a non-black respondent; you are much less likely to support Bush if your race is black compared to being non-black.
  
--

  + For a given state and race, a female respondent has $e^{-0.10}=0.91$ (a 9% decrease)  times the odds of supporting Bush as a male respondent. However, this effect is not actually statistically significant!



---
## 1988 elections analysis

- The state-level standard deviation is estimated at 0.41, so that the states do vary some, but not so much.

--

-   <div class="question">
We no longer have a term for residual standard deviation (residual standard error). Why is that?
</div>

--


- I expect that you will be able to interpret the corresponding confidence intervals.
  ```{r echo=F}
confint(model1)
```


---
## 1988 elections analysis

- Let’s fit a more sophisticated model that includes other relevant survey factors, such as 
  + region
  + prior vote history (note that this is a state-level predictor),
  + age, education, and the interaction between them.

--

- In `R`, we have
  ```{r results = 'hide',cache=T}
model2 <- glmer(bush ~ black + female + v_prev + edu_label:age_label +
                (1|state_label) + (1|region_label),
                family=binomial(link="logit"),data=polls_subset)
```

--

-   <div class="question">
Why do we have a rank deficient model?
</div>



---
## 1988 elections analysis

- Also, it looks like we have a convergence issue. This can happen when dealing with multilevel models. We have so many parameters to estimate from the interaction terms `edu_label:age_label` (16 actually), and it looks like that's causing a problem. 

--

- Could be that we have too many $\textrm{bush}_i = 1$ or $0$ values for certain combinations. You should check!

--

- Let's treat those as varying effects instead. That is,
.block[
.small[
$$
\begin{split}
\textrm{logit}\left(\Pr[\textrm{bush}_i = 1] \right) & = \beta_{0} + \gamma_{0m[i]}^{\textrm{region}} + \gamma_{0j[i]}^{\textrm{state}} + \gamma_{0k[i],l[i]}^{\textrm{age.edu}} \\
& + \beta_{\textrm{f}} \textrm{female}_{i} + \beta_{\textrm{b}} \textrm{black}_{i} + \beta_{\textrm{v_prev}} \textrm{v_prev}_{j[i]}; \\
\gamma_{0m} & \sim N(0, \sigma_{\textrm{region}}^2), \ \ \ \gamma_{0j} \sim N(0, \sigma_{\textrm{state}}^2), \ \ \ \gamma_{0k,l} \sim N(0, \sigma_{\textrm{age.edu}}^2).
\end{split}
$$
]
]

- In `R`, we have
  ```{r results = 'hide',cache=T}
model3 <- glmer(bush ~ black + female + v_prev + (1|state_label)
                + (1|region_label) + (1|edu_label:age_label),
                family=binomial(link="logit"),data=polls_subset)
```

--

- This seems to run fine; we are able to borrow information which helps.


---
## 1988 elections analysis

.small[
```{r echo=F}
summary(model3)
```
]



---
## 1988 elections analysis

- Remember that in the first model, the state-level standard deviation was estimated as 0.41. Looks like we are now able to separate that (for the most part) into state and region effects.

--

- Interpretation of results:
  + For a fixed state, education and age bracket, a non-black male respondent with zero prior average Republican vote share, has odds of $e^{-3.51}=0.03$ of supporting Bush (no one really has 0 value for `v_prev`).
  
--

  + For a fixed state, sex, education level, age bracket and zero prior average Republican vote share, a black respondent has $e^{-1.75}=0.17$ times (an 83% decrease) the odds of supporting Bush as a non-black respondent, which is about the same as before.
  
--

  + For each percentage point increase in prior average Republican vote share, residents of a given state, race, sex, education level  age bracket have $e^{0.07}=1.07$ times the odds of supporting Bush.


---
## 1988 elections analysis

- Due to the number of categories, the inference in the frequentist model is not entirely reliable as
  + it does not fully account for uncertainty in the estimated variance parameters, and 
  + it uses an approximation for inference. 
  
--

- We can fit the model under the Bayesian paradigm in the `brms` package, using mildly informative priors and quantify uncertainty based on the posterior samples.

--

- Windows users: install Rtools for windows, then the `rstan` package in R.

--

- Mac users: install Xcode, open it to accept the license agreement, then open R/RStudio and install the `rstan` package.

--

- .hlight[In-class analysis: move to the R script] [here](https://ids702-f21.olanrewajuakande.com/slides/Elections88.R).





---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




