---
title: "IDS 702: Module 6.6"
subtitle: "Propensity scores"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
#library(ggplot2)
#library(arm)
library(DiagrammeR)
library(MatchIt)
library(cobalt)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```



## Propensity scores

- The .hlight[propensity score] (ps) is defined as the conditional probability of receiving a treatment given pre-treatment covariates X.

--

- That is,
.block[
.small[
$$
e(X) = \mathbb{Pr}[W = 1 | X] = \mathbb{E}[W | X],
$$
]
]

  where $X = (X_1, \ldots, X_p)$ is the vector of $p$ covariates/predictors.

--

- Propensity score is a probability, analogous to a summary statistic.

--

- Propensity score has really nice properties which makes it desirable to use within our causal inference framework.


---
## Balancing property of propensity score

- .hlight[Property 1]. The propensity score e(X) balances the distribution of all $X$ between the treatment groups:
.block[
.small[
$$
W \perp X | e(X)
$$
]
]

--

- Equivalently,
.block[
.small[
$$
\mathbb{Pr}[W_i = 1 | X_i, e(X_i)] = \mathbb{Pr}[W_i = 1 | e(X_i)].
$$
]
]

--

- The propensity score is NOT the only .hlight[balancing score]. Generally, a balancing score $b(x)$ is a function of the covariates such that:
.block[
.small[
$$
W \perp X | b(X)
$$
]
]



---
## Remarks on the balancing property

- Rosenbaum and Rubin (1983) show that all balancing scores are a function of $e(X)$.

--

- If a subclass of units or a matched treatment-control pair are homogeneous in $e(X)$, then the treatment and control units have the same distribution of $X$.

--

- The balancing property is a statement on the distribution of $X$, NOT on assignment mechanism or potential outcomes.


---
## Propensity score: unconfoundedness

- .hlight[Property 2]. If $W$ is unconfounded given $X$, then $W$ is unconfounded given $e(X)$, i.e.,

--

- That is, if
.block[
.small[
$$
\{Y_i(0), Y_i(1)\} \perp W_i | X_i
$$
]
]

  holds, then
.block[
.small[
$$
\{Y_i(0), Y_i(1)\} \perp W_i | e(X_i),
$$
]
]

  also holds.

--
  
- Given a vector of covariates that ensure unconfoundedness, adjustment for differences in propensity scores removes all biases associated with differences in the covariates.
  
  
---
## Propensity score: unconfoundedness

- $e(X)$ can be viewed as a summary score of the observed covariates.

--

- This is great because causal inference can then be drawn through stratification,
matching, regression, etc. using the scalar $e(X)$ instead of
the high dimensional covariates.

--

- The propensity score balances the **observed covariates**, but does not generally balance **unobserved covariates**.

--

- In most observational studies, the propensity score e(X) is unknown and thus needs to be estimated.

--

- However, since we always observe $X$ and $W$, estimation can be done using models for binary outcomes.



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




