---
title: "IDS 702: Module 6.8"
subtitle: "The minimum wage analysis"
author: "Dr. Olanrewaju Michael Akande"
#date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
#library(ggplot2)
#library(arm)
library(DiagrammeR)
library(MatchIt)
library(cobalt)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```



## The minimum wage analysis

- Let's see how causal inference with propensity scores works using the minimum wage data.

--

- The data is in the file `MinimumWageData.csv` on Sakai.

  .small[
Variables    | Description
:------------- | :------------
NJ.PA | indicator for which state the restaurant is in (1 if NJ, 0 if PA)
EmploymentPre | measures employment for each restaurant before the minimum wage raise in NJ
EmploymentPost | measures employment for each restaurant after the minimum wage raise in NJ
WagePre | measures the hourly wage for each restaurant before the minimum wage raise
BurgerKing | indicator for Burger King
KFC | indicator for KFC
Roys | indicator for Roys
Wendys | indicator for Wendys
]


---
## The minimum wage analysis

```{r}
MinWage <- read.csv("data/MinimumWageData.csv",header=T,
                    colClasses=c("factor","numeric","numeric","numeric",
                                 "factor","factor","factor","factor"))
str(MinWage)
head(MinWage)
```


---
class: center, middle

# In-class analysis: move to the R script [here](https://ids702-f21.olanrewajuakande.com/slides/MinimumWage.R)



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




