---
title: "IDS 702 In-class analysis 2"
subtitle: "Beer consumption in Sao Paulo II"
#date: "August 25, 2020"
output: 
  tufte::tufte_html:
    tufte_variant: "envisioned"
    highlight: pygments
    css: "in-class-analysis.css"
    toc: true
link-citations: yes
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
knitr::opts_chunk$set(eval = FALSE)
```


**Due:** 1 hour after class ends



# Housekeeping
## Structure and format
You will work in your pre-assigned teams. Each team should submit ONLY ONE report for this exercise. You must write the names of all team members at the top of the report containing your responses. You all must do the work using one student's computer and R/RStudio. 

Have one team member open R/RStudio on their computer and share their screen with the other team members within the breakout room. At the top of the team report, write **"host"** in parenthesis besides this student's name. Have another team member be responsible for documenting the responses. At the top of the team report, write **"writer"** in parenthesis besides this student's name. 

**NOTE**: Generally, you will not be penalized for not taking on these roles many times during the semester. This is to simply ensure that you do switch the roles around a "decent number" of times within each team throughout the semester. That said, I will penalize any student who obviously dominates these roles over everyone else, so be sure to give other students an opportunity to do them.

## R/RStudio
You all should have R and RStudio installed on your computers by now. If you do not, first install the latest version of R here: https://cran.rstudio.com (remember to select the right installer for your operating system). Next, install the latest version of RStudio here: https://www.rstudio.com/products/rstudio/download/. Scroll down to the "Installers for Supported Platforms" section and find the right installer for your operating system.


## Gradescope
Gradescope will let you select your team mates when submitting, so make sure to do so. Only one person needs to submit the sheet on Gradescope. You can submit your document in the most common formats, but **pdf** files are preferred. Submit on Gradescope here: https://www.gradescope.com/courses/280392/assignments. Be sure to submit under the right assignment entry.



# Introduction
The purpose of this exercise is to give you additional practice working with multiple linear regression. We will continue with the Kaggle beer consumption dataset (https://www.kaggle.com/dongeorge/beer-consumption-sao-paulo/). You will demonstrate the impacts of some variables on beer consumption in a given region and the consumption forecast for certain scenarios. The data (sample) were collected in Sao Paulo, Brazil, in a university area, where there are some parties with groups of students from 18 to 28 years of age (average).

Kaggle is a great online community of data scientists. To learn more about Kaggle, follow this link: https://www.kaggle.com/getting-started/44916.



# The data
This is the same data from the last in-class exercise. so you should already have it saved locally. Just in case you do not, follow the instructions below.

Download the data (named `consumo_cerveja.csv`) from Sakai and save it locally to the same directory as your R markdown file. To find the data file on Sakai, go to Resources $\rightarrow$ Datasets $\rightarrow$ In-Class Analyses. Once you have downloaded the data file into the SAME folder as your R markdown file, load and clean the data by using the following R code.

```{marginfigure}
It is always a good idea to take a look at the first few rows of the raw file to see what the data looks like before loading the data. In this raw 'consumo_cerveja' file, you will notice that commas are actually used both as decimals and to separate the columns. Thus, you need to let R know by specifying the *sep* and *dec* options as in the code below.
```

```{r eval = FALSE}
beer <- read.csv("data/consumo_cerveja.csv",
                 stringsAsFactors = FALSE, sep = ",",
                 dec=",",nrows=365)
# rename the variables
beer$date <- beer$Data
beer$temp_median_c <- beer$Temperatura.Media..C.
beer$temp_min_c <- beer$Temperatura.Minima..C.
beer$temp_max_c <- beer$Temperatura.Maxima..C.
beer$precip_mm <- beer$Precipitacao..mm.
beer$weekend <- factor(beer$Final.de.Semana)
beer$beer_cons_liters <- as.numeric(beer$Consumo.de.cerveja..litros.)
beer <- beer[ , 8:ncol(beer)]
```

After renaming the variables using the code above, your data will be saved in the object `beer`, and the relevant variables plus their meanings are given in the table below:

Variable    | Description
:------------- | :------------ 
date | Date the data for each observation was recorded.
temp_median_c | Median temperature in $^0C$.
temp_min_c | Minimum temperature in $^0C$.
temp_max_c | Maximum temperature in $^0C$.
precip_mm  | Precipitation in $mm$.
weekend | Indicator variable for weekend: 1 = weekend, 0 = weekday.
beer_cons_liters | Beer consumption in liters.


# Exercises
Once again, fit a linear model for `beer_cons_liters` using `weekend`, `precip_mm`, and `temp_median_c` as your predictors.

1. Assess the fitted model. Describe in a few sentences whether or not you think the main regression assumptions are plausible based on residual plots (no need to include the plots). If you think any of the assumptions are violated, can you provide suggestions for how we might fix them?

2. How do your conclusions line up with the findings from your EDA in the last in-class exercise?

3. We will cover transformations later but make a histogram of `log(beer_cons_liters)`. Does that look more normal than `beer_cons_liters`? Fit a linear model for `log(beer_cons_liters)` using `weekend`, `precip_mm`, and `temp_median_c` as your predictors. Provide a 95% confidence interval for the effect of `weekend`. Interpret this interval in the context of the question. 

4. Assess the new fitted model. Describe in a few sentences whether or not you think the four main regression assumptions are plausible based on residual plots (no need to include the plots).

5. Overall, are there any potential limitations of this analysis? If yes, what are two potential limitations?



# Acknowledgement

This exercise is based on ideas proposed by [Sam Voisin](https://stat.duke.edu/people/sam-voisin).



